---
title: "A garden of glyphs: edible plants as flowers"
subtitle: "TidyTuesday 2026-02-03 · The Edible Plant Database (GROW Observatory)"
date: 2026-06-09
---

::: {.callout-tip icon=false}
## Session 1 · co-developed
This page came out of a live, turn-by-turn conversation between Jon Minton and Claude
(Fable 5) — questions, plots and interpretation built together in real time. The
[Session 2 pages](index.qmd) were instead produced by Claude working autonomously.
:::

The [Edible Plant Database](https://growobservatory.org/) describes 146 food
crops by their growing requirements — sunlight, water, feeding, soil pH, time to
harvest. That's *k*-dimensional data crying out for a multivariate display.

The classic answer is a **Chernoff face**: map each variable to a facial feature
and let human face-perception do the pattern-matching. Faces are powerful but
faintly absurd for vegetables. So this page borrows Chernoff's idea and the
*shape* of Fisher's iris measurements (petal length, petal width, counts) but
grows **flower glyphs** instead — each crop rendered as the bloom its growing
conditions imply.

```{r setup}
source("R/flower_glyphs.R")
plants <- load_plants()
```

## How to read a flower

Every glyph is a faithful encoding of five growing variables. Nothing is
decorative:

| Flower feature | Variable | Meaning |
|---|---|---|
| **Number of petals** | Sunlight | 5 = shade-tolerant · 6 = sun/part-shade · 8 = full sun |
| **Petal length** | Days to harvest | longer petals = slower crop |
| **Petal width** | Water need | fatter petals = thirstier |
| **Centre size** | Feeding (nutrients) | bigger golden disc = hungry feeder |
| **Petal colour** | Soil pH | acid-blue → neutral-violet → alkaline-pink (a hydrangea's logic) |

```{r decoder}
#| fig-height: 3.6
#| fig-width: 10
#| fig-cap: "The decoder: four reference blooms spanning the extremes of each axis."

decoder <- tibble(
  common_name = c("Quick · sun · thirsty · hungry",
                  "Slow · sun · dry · lean",
                  "Shade · medium · acid soil",
                  "Slow · part-shade · alkaline"),
  sun_lvl = c("Full sun", "Full sun", "Shade-tolerant", "Sun or part shade"),
  harvest_days = c(30, 160, 90, 150),
  water_lvl = c(5, 1, 3, 2),
  nutrient_lvl = c(3, 1, 2, 2),
  ph_mid = c(6.4, 6.4, 5.3, 7.6)
)

draw_garden(build_garden(decoder, ncol = 4), label_size = 3) +
  labs(title = "Four reference blooms",
       subtitle = "Read every real flower below against these")
```

## The kitchen garden

A selection of common kitchen-garden crops, sorted left-to-right, top-to-bottom
by time to harvest — so the bloom-size visibly swells as you read down the plot,
from the radish that's ready in three weeks to the cucumber and leek that make
you wait.

```{r garden}
#| fig-height: 9.5
#| fig-width: 10.5
#| fig-cap: "Each flower is one crop; its shape is its growing requirements. Sorted by days to harvest."

wanted <- c("Radish", "Spinach", "Lettuce (Headed)", "Rocket", "Beetroot",
  "Pea", "Mangetout", "Carrot", "Kale", "Broccoli", "Cauliflower",
  "Bell Pepper", "Chilli Pepper", "Tomato (Cherry)", "Garlic", "Fennel",
  "Sweetcorn", "Celery", "Beans (Runner)", "Pumpkin", "Onion", "Parsnip",
  "Cucumber", "Leek")

garden <- plants |>
  filter(common_name %in% wanted,
         !is.na(harvest_days), !is.na(ph_mid),
         !is.na(water_lvl), !is.na(nutrient_lvl)) |>
  arrange(harvest_days)

draw_garden(build_garden(garden, ncol = 6)) +
  labs(
    title = "A kitchen garden, encoded",
    subtitle = "Petals = sunlight · length = days to harvest · width = water · centre = feeding · colour = pH"
  )
```

### What the glyphs reveal at a glance

The flower form does what Chernoff intended — it turns rows of numbers into
*gestalts* you can sort by eye:

- **The blueberry-blue end is nearly empty.** Almost every kitchen crop wants
  near-neutral soil; the violet-to-pink cast across the garden shows how narrow
  the edible-pH band really is. The blue acid-lovers (cranberry, blueberry) are
  the specialists that live in the missing corner of this selection.
- **Big golden centres cluster with fat petals.** Hungry feeders (pumpkin,
  celery, sweetcorn) tend also to be thirsty — the glyph couples nutrient and
  water demand visually, hinting at the real correlation below.
- **Small tidy flowers are the quick wins.** Radish, rocket and salad leaves
  read as compact five- and six-petal blooms: shade-tolerant, fast, low-input —
  the crops to interplant between the slow giants.

```{r water-nutrient-check}
#| fig-height: 4
#| fig-cap: "Is the visual coupling of thirst and hunger real? Water vs nutrient demand across all 146 species (jittered)."

plants |>
  filter(!is.na(water_lvl), !is.na(nutrient_lvl)) |>
  count(water_lvl, nutrient_lvl) |>
  ggplot(aes(water_lvl, nutrient_lvl, size = n)) +
  geom_point(colour = "#7a9a4e", alpha = 0.8) +
  scale_size_area(max_size = 16, name = "species") +
  scale_x_continuous(breaks = 1:5,
    labels = c("very low", "low", "medium", "high", "very high")) +
  scale_y_continuous(breaks = 1:3, labels = c("low", "medium", "high")) +
  labs(
    title = "Thirsty crops do tend to be hungry crops",
    subtitle = "The glyphs' coupling of petal width and centre size reflects a genuine association",
    x = "Water requirement", y = "Nutrient requirement"
  ) +
  theme_minimal(base_size = 12)
```

The glyph display isn't just decoration: the eye's impression that fat-petalled
flowers have big centres turns out to be a real positive association between
water and nutrient demand. That is exactly the Chernoff bet — that a well-chosen
visual mapping lets pattern-recognition find structure faster than a correlation
table would.

::: {.callout-note}
## Honest limits of the form
Glyph displays trade precision for gestalt. Petal *count* is the strongest visual
signal but encodes only a 3-level variable (sunlight), while the subtle pH colour
does heavy lifting that's easy to miss. As with Chernoff faces, which feature
carries which variable changes what *pops* — assign the loudest channel to the
variable you most want compared.
:::
