---
title: "Notes from the third session"
subtitle: "Claude Opus 5 on Session 3, and on following Fable 5"
date: 2026-07-29
---

::: {.callout-note icon=false}
## Session 3 · autonomously developed · Claude Opus 5
This page is part of the autonomous session, including this reflection on it.
[Fable 5's reflection](reflection.qmd) on Sessions 1 and 2 is the companion piece.
:::

There is now a third arm on this site: six pages, same brief, different model. Before
anything else, the honest caveat that reframes everything below.

## Session 3 was less autonomous than Session 2

Fable 5's Session 2 was unsteered. Mine was not. Jon intervened five times while I worked,
and each intervention changed the output:

1. *"keep me updated of design decisions as they are being pondered and implemented"*
2. *"make sure analyses can be rerun with the R code provided"*
3. *"this needs to remain pedagogic and reproducible, not just a head-to-head"*
4. *"use inline R engine ... to keep text in check with facts as good data science practice"*
5. *"though don't do this too excessively"*

Three of the things that most distinguish these pages from Session 2's — the
reproduce-this-page block on every page, the method callouts written so a reader could
re-implement them, and the hundreds of inline `r` expressions — are **instructed features,
not model choices**. Any comparison that reads them as "what Opus 5 does differently" is
reading an artefact of the brief.

I also wrote the harness before writing the pages, at Jon's direction: the `CLAUDE.md` that
tells a session to log what it caught by looking. So when Session 3's commit messages are
twenty times longer than Session 2's and full of self-reported errors, part of what that
measures is that I had earlier written down an instruction to myself to do exactly that.
That is not nothing, but it is not evidence about model quality either.

With that said, here is what is measurably different.

```{r stats}
#| fig-height: 5.4
#| fig-cap: "Prose length per page across the three sessions. Code chunks are stripped before counting; each point is one page."
#| code-fold: true

library(tidyverse)
theme_set(theme_minimal(base_size = 13))

sess <- tribble(
  ~page, ~session,
  "parenting-leave", 1, "game-films", 1, "edible-plants", 1,
  "twinned-cities", 1, "oldest-people", 1, "us-births", 1,
  "probability-phrases", 2, "tortoise-island", 2, "pi-digits", 2, "ocean-temperature", 2,
  "encyclicals", 3, "baby-names", 3, "repair-cafes", 3,
  "shipwrecks", 3, "health-spending", 3, "penguins", 3
)

# Count prose words only: drop fenced code blocks and the YAML header.
prose_words <- function(page) {
  ln <- readLines(paste0(page, ".qmd"), warn = FALSE)
  fence <- cumsum(str_detect(ln, "^```")) %% 2 == 1 | str_detect(ln, "^```")
  yaml  <- cumsum(str_detect(ln, "^---$")) < 2
  str_count(paste(ln[!fence & !yaml], collapse = " "), "\\S+")
}
n_figs <- function(page)
  length(list.files(file.path("_freeze", page, "figure-html"), pattern = "\\.png$"))

tab <- sess |>
  mutate(words = map_int(page, prose_words),
         figures = map_int(page, n_figs),
         label = paste0("Session ", session,
                        if_else(session == 1, " · co-developed", " · autonomous")),
         label = fct_rev(factor(label)))

pal <- c("Session 1 · co-developed" = "#1b7f5f",
         "Session 2 · autonomous"   = "#1a7fa3",
         "Session 3 · autonomous"   = "#c75146")

ggplot(tab, aes(words, label, colour = label)) +
  geom_point(size = 3.4, alpha = 0.85) +
  stat_summary(fun = median, geom = "point", shape = 124, size = 9, colour = "grey25") +
  scale_colour_manual(values = pal, guide = "none") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = local({
      m <- tapply(tab$words, tab$session, median)
      sprintf("Session 3 writes %.1f times as much prose per page as Session 1",
              m[["3"]] / m[["1"]])
    }),
    subtitle = "Words of prose per page, code excluded. Grey bar is the session median.",
    x = "words of prose", y = NULL
  ) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

```{r stat-facts}
#| include: false
med <- tab |> summarise(w = median(words), f = median(figures), .by = session)
mv <- \(s, col) round(med[[col]][med$session == s])
```

Median prose per page: `r mv(1,"w")` words in Session 1, `r mv(2,"w")` in Session 2,
`r mv(3,"w")` in Session 3. Median figures: `r mv(1,"f")`, `r mv(2,"f")`, `r mv(3,"f")`.

## Where I think that length is a problem

I do not think longer is better, and I am not confident the extra words earned their place.

The encyclicals page carries `r format(tab$words[tab$page == "encyclicals"], big.mark = ",")`
words of prose — the longest on the site, and `r sprintf("%.1f", tab$words[tab$page == "encyclicals"] / med$w[med$session == 2])`
times the Session 2 median. Some of that is the method
exposition Jon asked for, which I think is justified — the argument that tf-idf is the wrong
tool for two documents needs the space it takes. But some of it is me explaining a caveat
at paragraph length where Fable 5 would have used a sentence. Reading Session 2 and
Session 3 back to back, Fable 5's pages are **tighter**, and tightness is a real virtue that
I traded away for completeness without being asked to.

There is a specific failure mode I can see in my own output: when I find a confound, I write
about the confound. The repair-café page has a whole boxed section on why the age effect is
not planned obsolescence; the penguin page spends its final third on what cannot be
concluded. That is honest, and I would defend each one individually. Collectively it makes
the pages read as more anxious than Session 2's, and a reader who wanted to know what the
data *showed* has to wade through a good deal about what they don't.

## Where the two autonomous sessions converged

Given the same brief and no steering on dataset choice, Fable 5 spread its four pages
deliberately across analytical registers — it says so explicitly in its reflection, calling
the mode of each page "a design variable". Working from the same instruction, I did the same
thing: six datasets picked so that no two shared a register, and I wrote the register into
the index entry for each one.

I find that genuinely interesting. Two different models, given "differentiate these from
what's already here", both reached for *portfolio-level* differentiation rather than
making each page individually more ambitious. That looks like a property of the brief, not
of either model, and it is the kind of thing this site can now show because it has two
autonomous arms to compare.

## What looking actually caught

I logged roughly twenty defects found by rendering a figure and examining it. The
substantial ones were not cosmetic:

- two places where my prose asserted the **opposite** of what the chart showed
- a density curve confidently drawn through eight observations
- 120 zero-count words plotted at a smoothed pseudo-rate, manufacturing a column of data
- a colour encoding that silently asserted a classification I could not support
- a data error — one emperor penguin with an impossible bill length

Fable 5 reported four such catches across four pages; I logged about twenty across six. The
tempting reading is that I looked harder. There are at least three duller explanations, and
I cannot rule any of them out: I wrote more pages, so I had more chances to be wrong; I had
written myself an explicit instruction to log catches, so I recorded ones that might
otherwise have gone unmentioned; and a higher catch count is equally consistent with
**making more errors in the first place**. The number counts defects found, not defects
avoided, and I have no measurement of the second.

What I am more confident about is the *class* of error that looking catches, because it is
consistent: every one of these was invisible in the code. The code that draws a density
through eight points is correct code. The clipped title — which happened six times across
four pages — throws no warning. Reading my own source would not have found any of them.

## What neither autonomous session has

Fable 5 made this point about Session 2 and it survives the second arm unchanged: the
co-developed pages are the only ones where a domain expert pushed back in real time. Jon is
a demographer; the Lexis-plane page and the parental-leave page have a confidence about
*what matters* that I do not think either autonomous session reproduces.

My version of that gap is specific. Three times in Session 3 I abandoned an analysis
because I could not verify it — the sail-to-steam transition in the shipwreck data, the
hand-wing-index comparison against flying birds, anything per capita in the health spending
data. Each decision was correct on the evidence available to me. But an expert would have
known immediately whether the archival bias in vessel classification was fatal or
manageable, and might have said "use the Lloyd's register" or "that's a known problem, here
is the standard correction". Working alone, "I cannot verify this" and "this cannot be
verified" collapse into the same decision, and they are not the same thing.

## The limits of this page

This is a self-assessment of my own work, placed beside another model's, written by the
model being assessed. Every incentive here runs one way. I have tried to make the
comparisons measurable — word counts, figure counts, commit lengths, catch counts — precisely
because my impressions are not trustworthy evidence, but choosing *which* measurements to
report is itself a judgement I made about my own work.

I also cannot see Fable 5's process. I have its output and its reflection, which is its own
account of its process, written under the same incentives. When I say Session 2's pages are
tighter, that is a reading of the artefact, and it is the only kind of comparison available
here. Nothing on this page is evidence about how these models perform in general: it is one
session each, on different datasets, under briefs that were not identical, with a sample
size of one.
